---
title: "Linear regression with one continuous explanatory variable"
source: Rmd
objectives:
- Use the ggplot2 package to explore the relationship between two continuous variables.
- Use the lm command to fit a simple linear regression with one continuous explanatory variable.
- Use the jtools package to interpret the model output.
- Use the jtools and ggplot2 packages to visualise the resulting model.
keypoints: 
- As a first check of suitability, construct a scatterplot between the continuous variables.
- Use `lm()` to fit the simple linear regression model.
- Use `summ()` to obtain parameter estimates for the model.
- Use `effect_plot()` to visualise the model. 
questions: 
- How can we assess whether two continuous variables may be suitable for simple linear regression?
- How is a simple linear regression model with one continuous explanatory variable fit in R?
- How can the parameters of this model be interpreted in R?
- How can this model be visualised in R?
teaching: 10
execises: 10
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
source("../bin/obtain_data.R")
knitr_fig_path("02-")
```

```{r load data, include=FALSE}
library(ggplot2)
library(jtools)
library(dplyr)
```

In this episode we will study linear regression with one continuous explanatory variable. As explained in the previous episode, the explanatory variable is required to have a linear relationship with the outcome variable. We can explore the relationship between two variables ahead of fitting a model using the `ggplot2` package.

Let us take `Weight` and `Height` of adults as an example. In the code below, we select adult participants with the command `filter(Age > 17)`. We then initiate a plotting object using `ggplot`. The filtered data is passed on by `data = .`. We select the variables of interest inside `aes()`. Finally, we create a scatterplot using `geom_point()`. Does the relationship look linear to you?

```{r Height vs Weight plot, warning = FALSE}
dat %>%
  filter(Age > 17) %>%
  ggplot(data = ., aes(x = Height, y = Weight)) +
  geom_point()
```

> ## Exercise  
> You have been asked to model the relationship between Urine Flow
> and Urine Volume in the NHANES data. In order to fit a simple 
> linear regression model, you first need to confirm 
> that the relationship between these
> variables appears linear. Use the ggplot2 package to create a plot,
> ensuring that it includes the following elements:  
> 1. Urine Flow (`UrineFlow1`) on the y-axis and Urine Volume
> (`UrineVol1`) on the x-axis, from the NHANES data.  
> 2. This data shown as a scatterplot.  
> 3. The y-axis labelled as "Urine Flow (mL/min)" and the x-axis labelled
> as "Urine Volume (mL)". 
>
> > ## Solution
> > 
> > ```{r, warning = FALSE}
> > ggplot(dat, aes(x = UrineVol1, y = UrineFlow1)) +
> >   geom_point(alpha = 0.4) +
> >   xlab("Urine Volume (mL)") + 
> >   ylab("Urine Flow (mL/min)")
> > ```
> {: .solution}
{: .challenge}



Since there is no abnormal shape to the scatterplot (e.g. curvature or multiple clusters), we can proceed with fitting our linear regression model. We do this with the `lm()` command. Within the command, the model is denoted by `outcome variable ~ explanatory variable`. Again, we subset our data using `filter()` and refer to this filtered data with `data = .`. See the code below. 

```{r fit Weigth vs Height}
Weight_Height_lm <- dat %>%
  filter(Age > 17) %>%
  lm(Weight ~ Height, data = .)
```

We will interpret our results through a summary table and through a plot. The summary table can be obtained using the `summ()` function from the `jtools` package. We provide the function with the name of our model (`Weight_Height_lm`). We can also specify that we want confidence intervals for our parameter estimates using `confint = TRUE`. Finally, we specify that we want estimates with three digits past the decimal with `digits = 3`.  

We will come to interpreting the `Model Fit` section in a later episode. For now, take a look at the parameter estimates at the bottom of the output. We see that the intercept (i.e. $\beta_0$) is estimated at -70.19 and the effect of `Height` (i.e. $\beta_1$) is estimated at 0.901. The model therefore predicts an average increase in `Weight` of 0.901 kg for every cm increase in `Height`.   

> ## Interpretation of the 95% CI and p-value of `Height`
> 
> If *prior to fitting our model* we were interested in testing the hypotheses 
> $H_0: \beta_1 > = 0$ vs $H_1: \beta \neq 0$, we could check the 95% CI 
> for `Height`. Recall that 95% of 95% CIs are expected to contain the true population 
> mean. Since this 95% CI does not contain 0, we would be fairly confident in rejecting 
> $H_0$ in favour of $H_1$. Alternatively, since the p-value is less than 0.05, 
> we could reject $H_0$ on those grounds. 
>
> Often the null hypothesis tested by `summ()`, $H_0: \beta_1 = 0$, is not very 
> interesting, as it is rare that we expect a variable to have an effect of 0. 
> Therefore, it is usually not necessary to pay great attention to the 
> p-value. On the other hand, the 95% CI can still be useful, as it provides an estimate
> of the *uncertainty* around our estimate for $\beta_1$. The narrower the 95% CI, 
> the more certain we are that our estimate for $\beta_1$ is close to the population 
> mean for $\beta_1$. 
{: .callout}

```{r summ Weight_Height_lm}
summ(Weight_Height_lm, confint = TRUE, digits = 3)
```

> ## Exercise  
> Now that you have confirmed that the relationship between Urine Flow
> and Urine Volume does not appear to deveate from linearity in the NHANES data, 
> you can proceed to fitting a simple linear regression model.  
>   
> 1. Using the `lm()` command, fit a simple linear regression of Urine Flow
> (`UrineFlow1`) as a function of Urine Volume (`UrineVol1`). 
> Name this `lm` object `UrineFlow_UrineVol_lm`.  
> 2. Using the `summ` function from the `jtools` package, answer the following questions:
>   
> A) What Urine Flow does the model predict, on average,
> for an individual with a Urine Volume of 0?  
> B) By how much is Urine Flow expected to change, on average, for
> a one-unit increase in Urine Volume?  
> C) Given these two values and the names of the response and explanatory
> variables, how can the general equation $E(y) = \beta_0 + {\beta}_1 
> \times x$ be adapted to represent your model? 
>
> > ## Solution
> > 
> > ```{r BPSysAve vs TotChol model}
> > UrineFlow_UrineVol_lm <- lm(UrineFlow1 ~ UrineVol1, data = dat)
> > 
> > summ(UrineFlow_UrineVol_lm, confint = TRUE, digits=3)
> > ```
> > 
> > A) 0.257 mL/min  
> > B) Increase by 0.006 mL/min  
> > C) $E(\text{Urine Flow}) = 0.257 + 0.006 \times \text{Urine Volume}$
> {: .solution}
{: .challenge}

We can also interpret the model through a line overlayed onto the previous scatterplot. We can obtain such a plot using `effect_plot()` from the `jtools` package. We provide the name of our model, followed by a specification of the explanatory variable of interest with `pred = Height`. Our current model has one explanatory variable, but in later lessons we will work with multiple explanatory variables. We also include a confidence interval around our line using `interval = TRUE` and include the original data using `plot.points = TRUE`. Finally, we specify a red color for our line using `colors = "red"`. 

```{r effect plot weight vs height}
effect_plot(Weight_Height_lm, pred = Height, plot.points = TRUE,
            interval = TRUE, colors = "red")
```


> ## Exercise  
> You have been asked to report on your simple linear regression model at 
> the next lab meeting. To help your colleagues interpret the model, you
> decide to produce a figure. Make this figure using the jtools package.
> Ensure that the x and y axes are correctly labelled.
>
> > ## Solution
> > ```{r}
> > effect_plot(UrineFlow_UrineVol_lm, pred = UrineVol1, plot.points = TRUE,
> >             interval = TRUE, colors = "red") +
> >   xlab("Urine Volume (mL)") + 
> >   ylab("Urine Flow (mL/min)")
> > ```
> {: .solution}
{: .challenge}


