---
title: "Linear regression with a two-level factor explanatory variable"
source: Rmd
objectives:
- Use the ggplot2 package to explore the relationship between a continuous variable and a two-level factor variable.
- Use the lm command to fit a simple linear regression with a two-level factor explanatory variable.
- Distinguish between the baseline and contrast levels of the categorical variable.
- Use the jtools package to interpret the model output.
- Use the jtools and ggplot2 packages to visualise the resulting model.
keypoints: 
- As a first exploration of the data, construct a violin plot between the two variables.
- Use `lm()` to fit the simple linear regression model.
- Use `summ()` to obtain parameter estimates for the model.
- The intercept estimates the mean in the outcome variable for the baseline group. The other parameter estimates the difference in the means in the outcome variable between the baseline and contrast group.
- Use `effect_plot()` to visualise the estimated means per group along with their 95% CIs. 
questions: 
- How can we explore the relationship between one continuous variable and one multi-level categorical variable prior to fitting a simple linear regression?
- How is a simple linear regression model with one two-level categorical explanatory variable fit in R?
- How does the use of the simple linear regression equation differ between the continuous and categorical explanatory variable cases?
- How can the parameters of this model interpreted in R?
- How can this model be visualised in R?
teaching: 10
execises: 10
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
source("../bin/obtain_data.R")
knitr_fig_path("03-")
library(ggplot2)
library(dplyr)
library(tidyr)
library(jtools)
library(Hmisc)
```

In this episode we will study linear regression with one two-level categorical variable. We can explore the relationship between two variables ahead of fitting a model using the `ggplot2` package.

Let us take `SmokeNow` and `TotChol` as an example. `SmokeNow` describes whether someone who has smoked > 100 cigarettes in their life is currently smoking. `TotChol` describes the total HDL cholesterol in someone's blood. In the code below, we first remove rows with missing values using `drop_na()` from the `tidyr` package. We then initiate a plotting object using `ggplot`. We select the variables of interest inside `aes()`. We then call for a violin plot using `geom_violin`. The shapes of the objects are representative of the distributions of `TotChol` in the two groups. We overlay the means and their 95% confidence intervals using `stat_summary()`. Finally, we change the axis labels using `xlab()` and `ylab()`.

```{r SmokeNow vs TotChol violin, warning = FALSE}
dat %>%
  drop_na(c("SmokeNow", "TotChol")) %>%
  ggplot(aes(x = SmokeNow, y = TotChol)) +
  geom_violin() +
  stat_summary(fun = "mean", size = 0.2) +
  stat_summary(fun.data = "mean_cl_normal", geom = "errorbar", width = 0.2) + 
  xlab("Participant has not given up smoking") +
  ylab("Total HDL Cholesterol")
  
```


> ## Exercise  
> You have been asked to model the relationship between average systolic
> blood pressure and physical activity in the NHANES data. Use the ggplot2
> package to create an exploratory plot, 
> ensuring that it includes the following elements:  
> 1. Average Systolic Blood Pressure (`BPSysAve`) on the y-axis and physical activity (`PhysActive`) on the x-axis, from the NHANES data.  
> 2. This data shown as a violin plot.  
> 3. The y-axis labelled as "Average Systolic Blood Pressure" and the x-axis labelled as "Physically Active".
>
> > ## Solution
> > 
> > ```{r, warning = FALSE}
> > dat %>%
> >   drop_na(c("PhysActive", "BPSysAve")) %>%
> >   ggplot(aes(x = PhysActive, y = BPSysAve)) +
> >   geom_violin() +
> >   stat_summary(fun = "mean", size = 0.2) +
> >   stat_summary(fun.data = "mean_cl_normal", geom = "errorbar", width = 0.2) +
> >   xlab("Physically Active")
> > ```
> {: .solution}
{: .challenge}

We proceed to fit a linear regression model using the `lm()` command, as we did in the previous episode. The model is then interpreted using `summ()`. 

Note that even though we are using the same equation of the simple linear regression model ($E(y) = \beta_0 + \beta_1 \times x_1$), our interpretation of the `summ()` output differs slightly from our interpretation in the previous episode. Recall that our categorical explanatory variable has two levels, `"No"` and `"Yes"`. One of these is taken by the model as the baseline (`"No"`), the other as the contrast (`"Yes"`). The first level alphabetically is chosen by R as the baseline, unless specified otherwise. 

The intercept in the `summ()` output is the estimated mean for the baseline, i.e. for participants that stopped smoking. The `SmokeNowYes` estimate is the estimated average difference in `TotChol` between participants that stopped smoking and participants that did not stop smoking. We can therefore write the equation for this model as:

$$E(\text{Total cholesterol}) = 5.053 + 0.008 \times x_1$$,

where $x_1 = 1$ if a participant has continued to smoke and 0 otherwise (i.e. the participant stopped smoking). 

```{r SmokeNow vs TotChol lm}
TotChol_SmokeNow_lm <- lm(formula = TotChol ~ SmokeNow, data = dat)

summ(TotChol_SmokeNow_lm, confint = TRUE, digits = 3)
```


> ## Exercise  
> 1. Using the `lm()` command, fit a simple linear regression of Average 
> Systolic Blood Pressure (`BPSysAve`)  
> as a function of physical activity (`PhysActive`). Name this `lm` 
> object `BPSysAve_PhysActive_lm`.  
> 2. Using the `summ()` function from the `jtools` package,
> answer the following questions:
>   
> A) What average systolic blood pressure does the model predict, 
> on average,
> for an individual who is characterised as not physically active?  
> B) By how much is average systolic blood pressure expected to change, 
> on average, for a physically active individual?  
> C) Given these two values and the names of the response and explanatory
> variables, how can the general equation $E(y) = \beta_0 + {\beta}_1 
> \times x$ be adapted to represent this model?
> 
> > ## Solution
> > 
> > ```{r}
> > BPSysAve_PhysActive_lm <- lm(formula = BPSysAve ~ PhysActive, data = dat)
> > 
> > summ(BPSysAve_PhysActive_lm, confint = TRUE, digits = 3)
> > ```
> > 
> > A) 122.892  
> > B) Decrease by 5.002  
> > C) $E(\text{BPSysAve}) = 122.892 - 5.002 \times \text{PhysActive}$, 
> > where $\text{PhysActive} = 0$ if an individual is not physically active
> > and $\text{PhysActive} = 1$ if an individual is physically active. 
> {: .solution}
{: .challenge}

Finally, we visually inspect the parameter estimates provided by our model. Again we can use `effect_plot()` from the `jtools` package. We include `jitter = 0.3` and `point.alpha = 0.1` so that points are spread out and so that multiple overlayed points create a darker colour, respectively. The resulting plot differs from the scatterplot obtained in the previous episode. Here, the plot shows the mean Total Cholesterol estimates for each level of `SmokeNow`, with their 95% confidence intervals. This allows us to see how different the means are predicted to be and within what range we can expect the true population means to fall.  

```{r SmokeNow vs TotChol effect_plot}
effect_plot(TotChol_SmokeNow_lm, pred = SmokeNow,
            plot.points = TRUE, jitter = 0.3, point.alpha = 0.1) +
  xlab("Participant has not given up smoking") +
  ylab("Total HDL Cholesterol")
```


> ## Exercise  
> Use the `jtools` package to visualise the model of `BPSysAve` as a 
> function of `PhysActive`.  
> Ensure that the x-axis is labelled as "Physically active" and the y-axis
> is labelled as "Average Systolic Blood Pressure".
> How does this plot relate to the output given by `summ`?
>
> > ## Solution
> > 
> > ```{r}
> > effect_plot(BPSysAve_PhysActive_lm, pred = PhysActive,
> > plot.points = TRUE, jitter = 0.3, point.alpha = 0.1) +
> >   xlab("Physically Active") +
> >   ylab("Average Systolic Blood Pressure")
> > ```
> > 
> > This plot shows the mean estimates for `BPSysAve` for the two groups, 
> > alongside their 95% confidence intervals. The mean estimates are 
> > represented by the `Intercept` for the non-physically active group and 
> > by `Intercept` + `PhysActiveYes` for the physically active group. 
> {: .solution}
{: .challenge}




