---
source: Rmd
title: "Making predictions from a simple linear regression model"
objectives:
  - "Calculate a prediction from a simple linear regression model using parameter estimates given by the model output."
  - "Use the predict function to generate predictions from a simple linear regression model."
keypoints:
- Predictions of the mean in the outcome variable can be manually calculated using the model's equation.
- Predictions of multiple means in the outcome variable alongside 95% CIs can be obtained using the `make_predictions()` function. 
questions:
  - How can predictions be manually obtained from a simple linear regression model?
  - How can R be used to obtain predictions from a simple linear regression model?
teaching: 10
exercises: 10
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
#source("../bin/obtain_data.R")
load("../bin/data.RData")
knitr_fig_path("04-")
library(ggplot2)
library(dplyr)
library(jtools)
```

One of the features of linear regression is prediction: a model presents predicted mean values for the outcome variable for any values of the explanatory variables. We have already seen this in the previous episodes through our `effect_plot()` outputs, which showed mean predicted responses as straight lines (episode 2) or individual points for levels of a categorical variable (episodes 3 & 4). Here, we will see how to obtain predicted values and the uncertainty surrounding them.

## Calculating predictions manually
First, we can calculate a predicted value manually. From the `summ()` output associated with our `Weight_Height_lm` model from episode 2, we can write the model as $E(\text{Weight}) = \beta_0 + \beta_1 \times \text{Height} = -70.194 + 0.901 \times \text{Height}$. The output can be found again below. If we take a height of 165 cm, then our model predicts an average weight of $-70.194 + 0.901 \times 165 = 78.471$ kg. 

```{r weight vs height model}
Weight_Height_lm <- dat %>%
  filter(Age > 17) %>%
  lm(formula = Weight ~ Height)

summ(Weight_Height_lm)
```

>## Exercise
>Given the `summ` output from our `BPSysAve_AgeMonths_lm` model, 
>the model can be described as
>
>$E(\text{BPSysAve}) = \beta_0 + \beta_1 \times \text{Age (months)} = 101.812 + 0.033 \times \text{Age (months)}$.  
>
>What level of average systolic blood pressure does the model predict, on average, 
>for an individual with an age of 480 months?
> > ## Solution
> > $101.812 + 0.033 * 480 = 117.652 \text{mmHg}$.
> {: .solution}
{: .challenge}

## Making predictions using `make_predictions()`
Using the `make_predictions()` function brings two advantages. First, when calculating multiple predictions, we are saved the effort of inserting multiple values into our model manually and doing the calculations. Secondly, `make_predictions()` returns 95% confidence intervals around the predictions, giving us a sense of the uncertainty around the predictions. 

To use `make_predictions()`, we need to create a `tibble` with the explanatory variable values for which we wish to have mean predictions from the model. We do this using the `tibble()` function. Note that the column name must correspond to the name of the explanatory variable in the model, i.e. `Height`. In the code below, we create a `tibble` with the values 150, 160, 170 and 180. We then provide `make_predictions()` with this `tibble`, alongside the model from which we wish to have predictions. By default, 95% confidence intervals are returned.

We see that the model predicts an average weight of 64.88 kg for an individual with a height of 150 cm, with a 95% confidence interval of [63.9kg, 65.9kg]. 

```{r Weight_Height_lm predict}
Heights <- tibble(Height = c(150, 160, 170, 180))

make_predictions(Weight_Height_lm, new_data = Heights)
```


>## Exercise
>1. Using the `make_predictions()` function, obtain the expected mean average systolic blood pressure levels
>predicted by the `BPSysAve_AgeMonths_lm` model for individuals with an age
> of 300, 400, 500 and 600 months.  
>2. Obtain 95% confidence intervals for these predictions.  
>3. How are these confidence intervals interpreted? 
> > ## Solution
> > ```{r}
> > BPSysAve_AgeMonths_lm <- dat %>% 
> >   filter(Age > 17) %>%
> >   lm(formula = BPSysAve ~ AgeMonths)
> > 
> > ages <- tibble(AgeMonths = c(300, 400, 500, 600))
> > 
> > make_predictions(BPSysAve_AgeMonths_lm, new_data = ages)
> > ```
> > Recall that 95% of 95% confidence intervals are expected to contain the 
> > population mean. 
> > Therefore, we can be fairly confident that the true population means lie 
> > somewhere between the bounds of the intervals, assuming that our model is good.
> {: .solution}
{: .challenge}

