---
source: Rmd
title: "Assessing simple linear regression model fit and assumptions"
objectives:
  - "Understand what is meant by model fit."
  - "Use the R squared value as a measure of model fit."
  - "Describe the assumptions of the simple linear regression model."
  - "Assess whether the assumptions of the simple linear regression model have been violated."
  - "Become aware of alternative regression method that can be used when the simple linear regression assumptions are violated."
keypoints:
  - "DEF"
questions:
  - "GHI"
teaching: 10
execises: 10
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("05-")
library(ggplot2)
library(dplyr)
library(jtools)
library(cowplot)
```

```{r load data, include=FALSE}
library(NHANES)
dat <- NHANES
```


Assumptions of the simple linear regression model:  
1. **Validity**: the model is appropriate for the research question. This sounds obvious, but it is easy to come to unreliable conclusions because of inappropriate model choice. Validity is assessed in three ways:  
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;A) Does the outcome variable reflect the *phenomenon of interest*? For example, it would not be appropriate to take our `Pulse` vs `PhysActive` model as representative of the effect of physical activity on general health.  
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;B) Does the model include *all relevant explanatory variables*? For example, we might decide that our model of `TotChol` vs `BMI` requires inclusion of the `SmokeNow` variable.  
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;C) Does the model generalise to our *case of interest*? For example, it would not be appropriate to model the effect of a *change* in physical activity on pulse using out `Pulse` vs `PhysActive` model. Neither would it be appropriate to take the model, which was constructed using people of all ages, as representative of the effect of physical activity on pulse in those aged 70+. In both examples, we would need a model constructed on different data to answer the question.  
2. **Representativeness**: the *sample* is representative of the *population*. More specifically, the individuals from which our sample is formed are representative of the population of interest. The exception to this requirement is that the sample distribution can differ from the population distribution in the explanatory variables included in the model. For example, let us assume that in the American population, 40% of individuals are physically active. In the NHANES data, ~56% of individuals are physically active. This discrepancy is dealt with by our `Pulse` vs `PhysActive` model, since `PhysActive` is an explanatory variable. However, if the majority of individuals in the NHANES data were over the age of 70, then our `Pulse` vs `PhysActive` model would not be representative of the American population. We would need to inlde `Age` as an explanatory variable to meet the representativeness assumption.  
3. **Linearity and additivity**: our outcome variable has a linear, additive relationship with the explanatory variables.  
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;The *linearity* component means that each explanatory variable needs to be modelled through a linear relationship with the outcome variable. For example, see plot **A** below. The relationship between `BPDiaAve` and `AgeMonths` is non-linear and our model `lm(BPDiaAve ~ AgeMonths , data=dat)` fails to capture this non-linear relationship. Adding a squared term to our model, designated by `I(AgeMonths^2)`, allows our model to capture the non-linear relationship (see plot **B**). Thus, the model `lm(BPDiaAve ~ AgeMonths + I(AgeMonths^2), data=dat)` does not violate the linearity assumption.  

```{r non-linearity example}
BPDiaAve_AgeMonths_lm <- lm(BPDiaAve ~ AgeMonths , data=dat)

p1 <- effect_plot(BPDiaAve_AgeMonths_lm, pred=AgeMonths, 
                  plot.points = TRUE, interval = TRUE,
                  colors = c("red")) +
  ylab("Combined diastolic blood pressure") +
  xlab("Age in Months") +
  ggtitle("Not a linear relationship") +
  theme_bw()

BPDiaAve_AgeMonthsSQ_lm <- lm(BPDiaAve ~ AgeMonths + I(AgeMonths^2), data=dat)

p2 <- effect_plot(BPDiaAve_AgeMonthsSQ_lm, pred=AgeMonths, 
                  plot.points = TRUE, interval = TRUE,
                  colors = c("red")) +
  ylab("Combined diastolic blood pressure") +
  xlab("Age in Months") +
  ggtitle("Non-linear relationship modelled \nusing an appropriate \nsimple linear regression model") +
  theme_bw()

plot_grid(p1, p2, labels=c("A)", "B)"))


```

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;The *additivity* component means that the effect of any explanatory variable on the outcome variable does not depend on another explanatory variable in the model. When this assumption is violated, it can be mitigated by including an interaction term in the model. We will cover interaction terms in the [multiple linear regression for public health lesson](https://carpentries-incubator.github.io/multiple-linear-regression-public-health/).  
4. **Independent errors**: the residuals must be independent of one another. This assumption is violated when observations are not a random sample of the population. For example, if we measure a individual's weights four times over the course of a year, then our data will contain four non-independent observations per individual. As a result, the residuals will also not be independent. This can be overcome using random effects, which we will cover in the [linear mixed effects models for public health] lesson] (https://carpentries-incubator.github.io/linear-mixed-models-public-health/).  
5. **Equal variance of errors**: the magnitude of variation in the residuals is not different across an explanatory variable. 

